Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-12 manuscript/main.tex
orthonormality proposition revised

%custom packages
 \usepackage{algorithm}
 \usepackage[noend]{algpseudocode}
 \usepackage{hyperref}
 \usepackage{enumerate}
 \usepackage{xcolor}
% \usepackage{url} % not crucial - just used below for the URL 
% \usepackage{bibunits}
\usepackage{subcaption}
% \usepackage{adjustbox}
\usepackage{verbatim} % to comment out

@@

A fundamental property of partial least squares   is that its derived bases are orthonormal. Our regularized estimates preserve this property, with respect to a modified inner product that incorporates the roughness penalty, defines as follows:
\begin{definition}[Roughness-sensitive inner product]\label{def:hybrid_inner_product_roughness}
	Given two hybrid predictors $W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$ and $W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, both elements of $\mathcal{H}$ as defined in Definition \ref{def:hybid_predictor}, and a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, the roughness-sensitive inner product between $W_1$ and $W_2$ is defined as:
	\begin{equation}		\label{eq: hybrid inner product_roughness}
		\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
		:=
		\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
		+
		\sum \limits_{k=1}^K \lambda_k \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
		+
		\mathbf{Z}_1^\top \mathbf{Z}_2.
	\end{equation}
\end{definition}
Based on this inner product, the following proposition states that the PLS component directions estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal.
\begin{proposition}\label{proposition: modified orthnormality of PLS components}\jongmin{currently revising}
	The estimated PLS components, $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, from the proposed PLS algorithm are mutually orthonormal with respect to the modified inner product $\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}$ in $\mathcal{H}$ that incorporates the roughness of the functions; i.e.,
	\begin{equation*}
		\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
		= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right\rangle_{L^2}
		= \mathbb{1}(l=j). 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.
Now let $\widehat{\boldsymbol{\rho}}_l=(\widehat{\rho}_{1l}, \ldots, \widehat{\rho}_{nl})^\top$ denote the $n$-dimensional vector whose elements consist of estimated $l$-th PLS scores ($l=1, \ldots, L)$ of $n$ observations.

($\widehat{\xi}_1, \widehat{\xi}_2, \ldots$)  based on the modified inner product  which now takes into account the roughness of the functions:
\begin{align*}
	\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
	&= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right \rangle_{L^2} \\
	&= \sum \limits_{k=1}^K \int_0^1 \widehat{\psi}_l^{(k)}(t_k) \widehat{\psi}_j^{(k)}(t_k) dt_k + \sum \limits_{r=1}^p \widehat{\theta}_{lr} \widehat{\theta}_{jr} + \sum_{k=1}^K \lambda_k \int_0^1 \ddot{\widehat{\psi}}_l^{(k)}(t_k) \ddot{\widehat{\psi}}_j^{(k)}(t_k) dt_k \\
	& = \mathbb{1}(l=j). 
\end{align*}

The next proposition states that the vectors of estimated PLS scores $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$ obtained from the proposed algorithm are mutually orthogonal.
\begin{proposition}	\label{proposition: orthnormality of PLS scores}
	The vectors of estimated PLS scores, $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$, obtained from the proposed PLS algorithm are mutually orthogonal in the sense that
	\begin{equation*}
		\widehat{\boldsymbol{\rho}}_l^\top \widehat{\boldsymbol{\rho}}_j= 0 \quad \textnormal{for} \quad \; l,j \in \{1, \ldots, L\}, \; l \ne j. 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.

>>>>

%custom packages
 \usepackage{algorithm}
 \usepackage[noend]{algpseudocode}
 \usepackage{hyperref}
 \usepackage{enumerate}
 \usepackage{xcolor}
 \usepackage{bbm}
% \usepackage{url} % not crucial - just used below for the URL 
% \usepackage{bibunits}
\usepackage{subcaption}
% \usepackage{adjustbox}
\usepackage{verbatim} % to comment out

@@

A fundamental property of partial least squares   is that its derived bases are orthonormal. Our regularized estimates preserve this property, with respect to a modified inner product that incorporates the roughness penalty, defines as follows:
\begin{definition}[Roughness-sensitive inner product]\label{def:hybrid_inner_product_roughness}
	Given two hybrid predictors $W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$ and $W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, both elements of $\mathcal{H}$ as defined in Definition \ref{def:hybid_predictor}, and a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, the roughness-sensitive inner product between $W_1$ and $W_2$ is defined as:
	\begin{equation}		\label{eq: hybrid inner product_roughness}
		\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
		:=
		\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
		+
		\sum \limits_{k=1}^K \lambda_k \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
		+
		\mathbf{Z}_1^\top \mathbf{Z}_2.
	\end{equation}
\end{definition}
Based on this inner product, the following proposition states that the PLS component directions estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal.
\begin{proposition}[Orthonormality of regularized PLS component directions]\label{proposition: modified orthnormality of PLS components}
	The PLS component directions $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, estimated via Proposition \ref{proposition:eigen_regul} with a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, are mutually orthonormal with respect to the inner product $\langle \cdot, \cdot \rangle_{\mathcal{H}, \Lambda}$. That is,
	\begin{equation*}
		\langle \widehat{\xi}_{l_1}, \widehat{\xi}_{l_2} \rangle_{\mathcal{H}, \Lambda}
		= \mathbbm{1}(l_1 = l_2), \quad l_1, l_2 = 1, \ldots, L.
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.

\jongmin{revising here}
Now let $\widehat{\boldsymbol{\rho}}_l=(\widehat{\rho}_{1l}, \ldots, \widehat{\rho}_{nl})^\top$ denote the $n$-dimensional vector whose elements consist of estimated $l$-th PLS scores ($l=1, \ldots, L)$ of $n$ observations. 
The next proposition states that the vectors of estimated PLS scores $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$ obtained from the proposed algorithm are mutually orthogonal.
\begin{proposition}	\label{proposition: orthnormality of PLS scores}
	The vectors of estimated PLS scores, $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$, obtained from the proposed PLS algorithm are mutually orthogonal in the sense that
	\begin{equation*}
		\widehat{\boldsymbol{\rho}}_l^\top \widehat{\boldsymbol{\rho}}_j= 0 \quad \textnormal{for} \quad \; l,j \in \{1, \ldots, L\}, \; l \ne j. 
	\end{equation*}
\end{proposition}
The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.
